Stage-Splitting a Modal Language

### 2014-07-24 paper/semantics.tex
increased exposition in intro

In this section, we describe \lang, a simple two-stage language
used as the basis for defining and analyzing stage splitting.
%The grammar and type system for \lang\ is given in
%\ref{fig:grammar,fig:statics}.
Every expression in \lang\ has a statically known {\em stage},
either \bbone~or \bbtwo, which defines when it is fully reduced.
%\footnote{Typing judgments and context variables are annotated with stages.}
%
% in addition to a type.
% Just as a type describes \emph{which} values an expression reduces to,
%the expression's stage tells us \emph{when} it will be fully reduced. 
% FIXME why?
%The reason for a two-stage design (as opposed to \cite{davies96}, which allows an infinite number of stages) will be discussed later.
Stage-\bbone\ and stage-\bbtwo\ types include products, sums, functions, and base types.
In addition to these types, stage-\bbone\ expressions may also be of type $\fut A$,
representing an enclosed stage-\bbtwo\ expression of type $A$.
(The type system of \lang\ is adopted directly from that of the modal language \lamCircle\,\cite{davies96}, but is restricted to two stages for simplicity.) $\fut A$ terms must be treated opaquely by stage-\bbone\ code
as they cannot be reduced (to values with type $A$) until stage \bbtwo. Like in \lamCircle\, there is no reciprocal modality.

Although there is a strict one-way dependence between the two stages at the type
level, stage-\bbone\ and stage-\bbtwo\ terms are defined mutually recursively:
the $\next$ construct embeds stage-\bbtwo\ expressions into stage \bbone,
whereas $\prev$ embeds stage-\bbone\ expressions into stage \bbtwo.
The $\next$ and $\prev$ constructs are the only expressions that interact with the stage of a term,
and we surround their arguments with braces in \lang\ syntax to clearly indicate stage boundaries within a program.

>>>>

In this section, we describe \lang, a simple two-stage language
used as the basis for defining and analyzing stage splitting.

%The grammar and type system for \lang\ is given in
%\ref{fig:grammar,fig:statics}.
Every expression in \lang\ has both a type and a {\em stage}, either \bbone~or \bbtwo.
Just as a type describes \emph{what} sort of value an expression reduces to, the stage tells us \emph{when} an expression will be fully reduced.  
Within each stage, \lang\ reduces to the simply-typed lambda calculus augmented with products, sums, general recursion, and base types
\footnote{It needn't be the case that we choose the same set of features at both levels, but this does simplify the exposition.}.

In addition to these types, stage-\bbone\ expressions may also be of type $\fut A$,
representing an enclosed stage-\bbtwo\ expression of type $A$. $\fut A$ terms must be treated opaquely by stage-\bbone\ code
as they cannot be reduced (to values with type $A$) until stage \bbtwo.
While $\fut$ allows use to embed stage-\bbtwo\ types within stage-\bbtwo\ types, 
there is no way to embed a stage-\bbone\ type within a stage-\bbtwo\ type.

Despite this one-way dependence between the two stages at the type level, 
stage-\bbone\ and stage-\bbtwo\ terms are defined mutually recursively:
the $\next$ construct embeds stage-\bbtwo\ expressions into stage \bbone,
whereas $\prev$ embeds stage-\bbone\ expressions into stage \bbtwo.
The $\next$ and $\prev$ constructs are the only expressions that interact with the stage of a term,
and we surround their arguments with braces in \lang\ syntax to clearly indicate stage boundaries within a program.
